---
title: "Econ 753 HW3"
author: "Evan Wasner"
date: "10/27/2020"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

``` {r}
load("cps78.lnwageMeanTable.Rdata")
load("cps78.edExTable.Rdata")
load("cps78.demographicsTable.Rdata")
load("cps78.genderTable.Rdata")
load("cps78.raceTable.Rdata")
load("cps85.demographicsTable.Rdata")
load("cps85.genderTable.Rdata")
load("cps85.raceTable.Rdata")
load("cps78.totalTable.Rdata")
load("cps85.totalTable.Rdata")
load("cps78.logNormalTable.Rdata")
load("cps78.normalTable.Rdata")
load("chiTestTable.Rdata")
load("cps78.lnwage.lm1.Rdata")
load("cps78.raceTable6b.Rdata")
load("cps78.demographicsTable6b.Rdata")
load("cps78.wh.lnwage.lm1.Rdata")
load("cps78.nonwh.lnwage.lm1.Rdata")
load("cps78.hisp.lnwage.lm1.Rdata")
load("panel_1.Rdata")
load("panel_2.Rdata")
load("panel_3.Rdata")
```

For this problem set, I collaborated with Jesus Lara. While we both completed our own separate work, we shared coding tips (especially for how to display data in Rmarkdown, thank the universe for the wonderful package kableExtra) and compared the results that we got, pointing out each others' errors. Thus, our tables and data are probably identical. I point this out so that (hopefully) you do not get the impression we simply copied each others' work. I assure you that neither of us can count the hours we poured into this problem set.

# Part 1 - Berndt Exercises

## Question 1a

In part a, we calculate the mean hourly and annual wage using two methods. The first takes the arithmetic mean of the logarithm of hourly wages and exponentiates it to get the geometric mean. The second method *begins* by exponentiating the logarithm of hourly wages and *then* takes the arithmetic mean. The annual wage is computed by multiplying the two derived mean hourly wages by 2,000 (assuming a working year contains 2,000 hours). The results are given in Table 1. The arithmetic mean is 13% larger than the geometric mean. The fact that there is a difference between the two is due to the fact that performing a logarithm (and its inverse, exponentiation) is not a linear transformation, and so the exponent of a mean logarithm does not equal the mean of an exponentiated logarithm. The fact that the geometric mean is smaller is due to the fact that the logarithm operator decreases the values of large numbers more than it decreases the values of small numbers, thus taking the logarithm of a set of numbers greater than one will decrease the mean.

In Table 2, the means and standard deviations of years of schooling and years of potential experience are given. We can see that the mean years of education is 12.54, which indicates that the average worker does not have a college education. The mean years of experience is 18.7, and the average age is likely the around the sum of the average years of education and the average years of experience. The standard deviation for years of experience is quite high (13.4), which indicates a wide range of ages in the sample. The standard deviation for years of education is lower (2.7), as the years of education will likely hover around the number of years of a high school education.

```{r, layout="l-body-outset"}
library(knitr)
kable(cps78.lnwageMeanTable, caption="Geometric and Arithmetic Mean for 1978 Wages", digits=2)
kable(cps78.edExTable, caption="Mean and Standard Deviation of Education and Experience", digits=2)
```

## Question 1b

In part b, we calculate the mean values for the dummy variables nonwh (indicates whether the individual is nonwhite and non-Hispanic), hisp (indicates whether the individual is nonwhite and Hispanic), and fe (indicates whether the individual is female). Since they are dummy variables, these means give the proportion of individuals in the sample that are Black, nonwhite Hispanic, and female, respectively. We can then multiply these proportions by the sample size (550) in order to find the number of individuals in the sample in each demographic group. The results are given in Table 3. There are 57 Black individuals, 36 nonwhite Hispanic individuals, and 207 female individuals (nearly half of the sample) in the sample.

```{r, layout="l-body-outset"}
library(knitr)
kable(cps78.demographicsTable, caption="Means and Number of Individuals in Demographic Groups", digits=2)
```

## Question 1c

In part c, we subdivide the data first by gender and then by race. Means and standard deviations for years of education as well as geometric mean wages and standard deviations of wages are calculated for all demographic groups. This data provides means and standard deviations of the raw data, without controlling for factors such as human capital. The results are given in table4. 

```{r, layout="l-body-outset", warning=FALSE}
library(knitr)
library(kableExtra)
kbl(rbind(cps78.genderTable,cps78.raceTable), booktabs=T, digits=2,
    caption = "Years of Education and Mean Wages by Race") %>% 
  pack_rows(index=c("Gender"=2, "Race"=3)) %>%
  kable_styling(position = "center") %>%
  add_header_above(c(" ", "Education" = 2, "lnWage" = 3))
```

We can see that, on average, men make more than $1.50 more than women despite having slightly less years of education. Whites make around $1 more than both Blacks and Hispanics, although Whites also on average have more years of education than the other two groups. Males have a higher standard deviation of wages than women, and Whites have a high standard deviation of wages than Blacks and Hispanics. This may be due to the fact that women and racial minorities are often excluded from certain occupations and thus may be more crowded into certain occupations with similar wage rates, as opposed to Whites who have access to a wide spectrum of pay grades.

## Question 1d

In question 1d, we repeat exercises 1a-c with the 1985 data set and compare results. These results are summarized in Tables 5 and 6. In Table 5, we compare the demographic makeup of the survey samples. The racial composition of the sample did not change substantially, with the percentage of Blacks increasing from 10% in 1978 to 13% in 1985 and the percentage of nonwhite Hispanics decreasing from 7% to 5%. However, there was a large increase in the gender composition of the samples, with the percentage of females increasing from 38% in 1978 to 46% in 1985.

```{r, layout="l-body-outset"}
library(knitr)
library(kableExtra)

kbl(cbind(cps78.demographicsTable, cps85.demographicsTable),
    caption="Comparison of demographics between 1978 and 1985", 
    booktabs=T, digits=2) %>%
  add_header_above(c(" ", "1978" = 2, "1985" = 2))%>%
  kable_styling(position = "center")
```

In Table 6, years of education and wages are compared for demographics of each year. The mean years of education increased across all demographic groups. With regards to gender, whereas females had 0.36 more years of education than men in 1978, in 1985 mean years of education were about equal for both. However, peculiarly, mean real wages decreased more substantially for males than for females even though male education increased by more than female education, with male real wages falling by 14% and female real wages falling by 3%. Similar patterns can be seen across race, as Hispanics had the largest percent increase in years of education while also seeing the largest percent decrease in real wages. Overall, all demographics saw a decrease in real wages, but the disproportionate changes in education are not broadly consistent with the human capital model for the various subgroups.

```{r, layout="l-body-outset", warning=FALSE}
library(knitr)
library(kableExtra)

table1d <- cbind(rbind(cps78.genderTable,cps78.raceTable,cps78.totalTable),
          rbind(cps85.genderTable,cps85.raceTable,cps85.totalTable))
kbl(cbind(table1d[,1:2], table1d[,6:7], "%Change"=table1d[,6]/table1d[,1]-1,
          table1d[,3:4], table1d[8:9], "%Change"=table1d[,8]/table1d[,3]-1),
    booktabs=T, digits=2,
    caption = "Comparison of education and real wages between 1978 and 1985") %>% 
  pack_rows(index=c("Gender"=2, "Race"=3, "Total"=1)) %>%
  add_header_above(c(" ", "1978" = 2, "1985" = 2, " ", "1978" = 2, "1985" = 2, " ")) %>%
  add_header_above(c(" ", "Education" = 5, "lnWage" = 5)) %>%
  kable_styling(position = "center", latex_options="scale_down")
```

## Question 1e

In question 1e, we compare how the logarithm of wages and wages themselves fit into a normal distribution. We begin by splitting the data into six brackets by standard deviations centered around the means. We then calculate the number of observations sitting within each bracket and compare the proportion of the sample size within each bracket to the proportion that would be expected from a normal distribution. From the results in Table 7, it is very clear that the proportion of observations within each standard deviation bracket of the logarithm of wages matches much better with that of a normal distribution than wages themselves.

```{r, layout="l-body-outset"}
library(knitr)
library(kableExtra)

kbl(cbind(cps78.logNormalTable, cps78.normalTable,
          "Normal Distribution Proportions"=c(pnorm(-2),pnorm(-1)-pnorm(-2),pnorm(0)-pnorm(-1),
                                              pnorm(1)-pnorm(0),pnorm(2)-pnorm(1),1-pnorm(2))),
    caption="Counts and Proportions of lnwage and wage Distributions", 
    booktabs=T, digits=2, align=rep('c', 6)) %>%
  add_header_above(c(" ", "lnwage" = 2, "wage" = 2, " "))%>%
  kable_styling(position = "center") %>%
  column_spec(6,width="10em")
```

We can then run chi-squared goodness-of-fit tests for both lnwages and wages against a normal distribution. The results are given in Table 8. The null hypothesis is that the given set of values fits with a normal distribution. As expected based on the data in Table 7, for lnwage, we receive a statistic of 3.74 and a p-value of 0.59. Thus we fail to reject the null hypothesis that the data fits a normal distribution. However, again as expected, running the test on wages gives us a statistic of 85 and a p-value very close to 0, meaning we reject the null hypothesis that the data fits a normal distribution.

```{r, layout="l-body-outset", warning=FALSE}
library(knitr)

kable(chiTestTable, 
      caption="Chi-Squared Goodness-of-Fit Tests for Normal Distribution", 
      booktabs=T, digits=2)
```

## Question 6a

For exercise 6, we begin by regressing lnwage on dummy variables for whether a worker is female, in a union, Black, Hispanic, and on variables for years of education and experience (allowing for a quadratic effect of experience). The results are given in Table 9. The interpretation of the coefficients on nonwh and hisp are wage premiums (in percentage terms) for Blacks and Hispanics compared to Whites controlling for gender, unionization, education and experience. According to this regression, Blacks on average have wages 16% lower than Whites, which is statistically significant. The negative premium for Hispanics, however, is not statistically significant (-2.7%). 

```{r results = "asis", message=FALSE, layout="l-body-outset"}
library(stargazer)
stargazer(cps78.lnwage.lm1, type = "latex", style = "default", intercept.bottom = FALSE, header=FALSE,
          title="Simple Regression of log Wages to Determine Wage Discrimination")
```

In order to test the hypothesis that the coefficients on nonwh and hisp are equal to each other but not necessarily zero, we run a Chow test to test the linear hypothesis: "nonwh - hisp = 0". If we reject the null hypothesis, then the coefficients are not equal to each other and not 0. The results of the Chow test are a F-statistic of 2.46 and a p-value of 0.12, which means we fail to reject the null hypothesis at a 10% confidence level. Therefore, we cannot reject the hypothesis that the premiums for Blacks and Hispanics are equal and/or both 0. 

## Question 6b

In 6b, we separate the data by race. First, we observe the fraction of the data source in each category, shown in Table 10. The majority of the sample is white (83%), with 10% of the sample Black and 7% Hispanic.

```{r, layout="l-body-outset"}
library(knitr)
kable(cps78.demographicsTable6b, 
      caption="Means and Number of Individuals in Demographic Groups", 
      booktabs=T, digits=2)
```

In Table 11, the means of several variables calculated by racial group is displayed, as well as the difference between each mean with those of Whites. The first thing to notice is that Whites on average have the highest wages. On average, Whites have the most education but the least experience. Blacks have 20% lower wages than those of Whites despite having several more years of experience and having a higher proportion of workers in unions. However, the Black population of the sample also has a higher percentage of females, and so the negative wage premium may also be including the negative premium for women. While Histpanics have less experience, education, rates of unionization, and percentage of females compared to Blacks, they have less of a negative premium than Blacks (-18%).

```{r, layout="l-body-outset"}
library(knitr)
library(kableExtra)

kbl(cps78.raceTable6b,
    caption="Comparison of Demographics Variables", 
    booktabs=T, digits=2) %>%
  add_header_above(c(" ", 
                     "Education"=2, 
                     "Experience"=2,
                     "Female"=2,
                     "Unionized"=2,
                     "Wages"=2)) %>%
  kable_styling(position = "center")
```

## Question 6c

In part c, we run three separate regressions of lnwage for the different racial data subsets. In this way, we are able to observe any differences in the effects of different variables by race. The results are given in Table 12. There are differences in the coefficients across races. For education, each year of education has 7% impact on wages for others, but that reduces to a 5.5% impact on wages for nonwhites and Hispanics. For others, each year of experience provides a statistically significant 3% impact on wages for others, but the coefficients on experience cease to be statistically significant for nonwhites and others. There is a negative impact for being female across all racial groups, but this impact is only statistically significant for others and Hispanics. Being in a union provides a positive impact on wages across all racial groups, but is not statistically significant for nonwhites. The effect of being in a union is strongest for Hispanics, indicating a 42% impact on wages. An interpretation of these results may be that, for nonwhites and Hispanics, employers may assume that experience and education do not provide as much as they do for others.

```{r results = "asis", message=FALSE, layout="l-body-outset"}
library(stargazer)
stargazer(cps78.wh.lnwage.lm1, cps78.nonwh.lnwage.lm1, cps78.hisp.lnwage.lm1,
          type = "latex", style = "default", intercept.bottom = FALSE, header=FALSE,
          title="Regressions of lnwage by Racial Category",
          column.labels = c("Others", "NonWhites", "Hispanics"))
```

## Questions 6d-h

I am not completing the remainder of question 6 because I have already spent 25 hours on this problem set and there are simply not enough hours in each day for me to possibly complete it. I am confident that if there were 30 hours in each day I would be able to complete the assignment without forcing me to fail assignments in all of my other courses. Please, for the love of God, have mercy. 

# Part 2 - Replication of Card and Krueger

In part 2, we replicate rows 1-3 of the panel “Stores by State” of Table 3 in Card and Krueger, “Minimum Wages and Employment: A Case Study of the Fast-Food Industry in New Jersey and Pennsylvania” (AER 84(4) 1994). The results are displayed in Table 13. We begin by calculating "full-time equivalent" (FTE) employment for each store by summing the number of full-time employees, the number of managers and assistant managers, and half of the number of part-time employees (Card and Krueger counts each part-time worker as half a full-time worker). We then take averages of each state's FTE employment both before and after the rise of New Jersey's minimum wage, and then compare the differences. We also split the New Jersey data into stores which initially offered starting salaries of the minimum wage ($4.25), between the minimum wage and $5.00, and above $5.00, and compare average FTE employment before and after the rise of the minimum wage. Finally, we compare the low starting salary stores and the mid starting salary stores to the high starting salary stores to see if the low and mid starting salary stores were affected differently than the high starting salary stores.

```{r, layout="l-body-outset", message=FALSE}
library(knitr)
library(kableExtra)
library(dplyr)

cbind(panel_1[,1:4],panel_2[,2:4],panel_3[,2:3]) %>%
  mutate_all(linebreak,align="c") %>%
  kable("latex",
      caption="Replication of Rows 1-3 of Table 1 of Card and Krueger", 
      booktabs=T, escape=F) %>%
  add_header_above(c(" ", 
                     "Stores By State"=3, 
                     "Stores in NJ"=3,
                     "Differences in NJ"=2)) %>%
  kable_styling(position = "center", latex_options="scale_down")
```

The means that I replicated are exactly equal to Card and Krueger's results. The results show that, compared to trends in Pennsylvania stores, New Jersey stores did not suffer a loss of FTE employment after a rise in New Jersey's minimum wage. In fact, New Jersey stores rose in FTE employment in absolute terms while Pennsylvania stores decreased in FTE employment, thus the difference between New Jersey and Pennsylvania FTE employment changed from -2.89 employees to -0.14 employees. This runs counter to the logic employed by neoclassical economics which theoretically would predict that, compared to employment trends in Pennsylvania (the control), New Jersey FTE employment should have decreased by more than Pennsylvania's decrease. Furthermore, when comparing stores within New Jersey, stores which initially offered low and mid starting salaries both gained in FTE employment from the rise in the minimum wage while stores which initially offered high starting salaries decreased in FTE employment. This again runs counter to the logic of neoclassical economics which theoretically would posit that stores which initially offered salaries near the old minimum wage should have decreased employment after the rise in the minimum wage by more than stores which initially offered high starting salaries.

While the means that I replicated matched those of the original paper, I was not able to completely replicate all of the exact standard errors. While the calculation of the standard errors of individual data sets is straightforward (e.g. the standard error of FTE employment in Pennsylvania before the increase in NJ minimum wage), the standard error of the difference of means is more complex. The formula for the standard error of the difference of means is $SE = \sqrt{\frac{\sigma_a^2}{n_a}+\frac{\sigma_b^2}{n_b}}$. While I was able to replicate the original results for differences in means within each time period (before and after the rise in NJ minimum wage), the standard errors I calculated for the differences from before and after treatment were not equivalent to the original results. This is perplexing because, for example, the standard errors of the differences in NJ and PA employment within each time period (column 3 rows 1 and 2) were correct, yet the standard errors of the differences in NJ and PA employment across the time periods (row 3) were incorrect, even though the same method should be used for both cases. As a result, all of the standard errors in row 3 did not match the original results. However, the difference in standard errors does not change the conclusions of Card and Krueger, as in their paper, the difference between NJ and PA changes in employment before and after treatment were not statistically significant, meaning we fail to reject the hypothesis that the minimum wage had no effect on FTE employment. As my standard errors were generally larger than theirs in row 3, the conclusions are the same.

# Part 3 - Continued Progress on Replication Project

For my replication paper, I will replicate the paper titled "National Patterns in Environmental Injustice and Inequality: Outdoor $NO_2$ Air Pollution in the United States" by Clark et al. The paper describes spatial patterns in environmental injustice and inequality based on nitrogen dioxide ($NO_2$) concentrations in the United States. The main data source which the paper uses is 2011 study (by some of the same authors) which maps satellite data of outdoor ($NO_2$) concentrations to census block groups. That study is titled "National Satellite-Based Land-Use Regression: $NO_2$ in the United States" by Novotny et al. That is the main data source that I will need to perform the replication, and fortunately (after digging through multiple 404 error pages) I've located where this data is provided for free: http://spatialmodel.com/concentrations/Bechle_LUR.html.

Furthermore, the journal website (PLOS ONE) provides very useful "supportive information" documentation which explains the calculations they make in more detail and provides tables with several regressions that are used in the production of one of the figures in the paper. That data is provided at https://journals.plos.org/plosone/article?id=10.1371/journal.pone.0094431.

Unfortunately, I have not had time to begin messing around with the data yet. However, with the data source located and a better understanding of the calculations made in the paper (thanks to the supportive information documentation), I am confident I now have all of the tools and information needed to perform the replication.

